Constructing Sailing Match Race Schedules: Round Robin Pairing Lists

### 2015-04-08 paper.tex
Spacing

This work describes a round robin competition format that arises from sailing, known as match
racing. Competitors, i.e.\ \emph{skippers}, compete against each other in a series of matches taking
place in rounds, known as \emph{flights}.  A match is composed of two skippers, with each skipper in
a boat on his own. Skippers in a match set off together, one on the port side, the other starboard,
and first home wins that match. Criteria published by the International Sailing Federation (ISAF)
dictate what makes a legal match race schedule and what should be optimised. This is a rich set of
criteria, 13 in all, and as far as we know, all published schedules have been produced by hand. This
is a daunting task. A close inspection of these schedules reveals that most are illegal, violating
many of the match race scheduling criteria, many schedules are missing, and some that are legal are
far from optimal.

@@

@@

We now describe the constraint models and processes used in each stage. In the text below we assume
there are $n$ skippers and $b$ boats, with $m = b/2$ matches in a flight. In total there are $t =
n(n-1)/2$ matches and $f = \lceil n(n-1)/m  \rceil$ flights. We use the schedule in
Table~\ref{tab1}, for 7 skippers and 6 boats, to help illustrate the constraint models. Our
descriptions of the models omit some minor details, however the models are available to
download\footnote{?? give us a link to the models}.

@@

\paragraph{Modeling skippers:} The first thing we model, in the first half of
Figure~\ref{model:stage1}, is a skipper. Each skipper $\sigma$ has a temporal view of his schedule
in the array of constrained integer variables $\timeSlot[0]$ to $\timeSlot[{t-1}]$, and a state view
in the array of constraint integer variables $\state[0]$ to $\state[{f-1}]$. The state of this
skipper in flight $i$ is found in $\state[i]$, and this corresponds to timeslots $\timeSlot[{i \cdot
m}]$ to $\timeSlot[{m(i+1)-1}]$.

@@

In \eqref{var:timeSlot} the array $\timeSlot$ is defined and in \eqref{var:state}, $\state$ is
defined. If variable $\timeSlot[{i}] = k$ and $k \geq 0$ then this skipper is in a match with
skipper~$k$ at time~$i$. Variable $\state[{i}]$ gives the state of this skipper in flight~$i$. The
cases in \eqref{C1} state that a skipper can be in a match in the first time slot in the flight, or in
the middle of the flight\footnote{i.e.\ not first and not last.}, or in the last time
slot of the flight. Alternatively, if all time slots in a flight are equal to $-1$ then the skipper
is in a bye, and if all time slots in a flight are equal to $-2$ then the skipper has finished all
his matches.

We must then ensure that each skipper $\sigma$ is in a match with all other skippers $\{0, \nldots,
n-1\} \setminus \{\sigma\}$. This is \eqref{C2}, which may be implemented by imposing a global
cardinality constraint \cite{globCard} on the array $\timeSlot$.

\paragraph{State transitions:} The $\state$ variables are then used to impose match race criterion~4
(if last in flight $i$ then not first in flight $i+1$), criterion~11 (if last in flight $i$ then not
in a bye in flight $i+1$) and criterion~12 (if flight $i$ is a bye then flight $i+1$ is not first).
These criteria are imposed by \eqref{C3}, which is a deterministic finite automaton (DFA) constraint
\cite{Pesant04}. The transitions for the DFA are shown pictorially in Figure~\ref{skipper5}: arcs
represent a possible transition from one state to another. The DFA is encoded as a set of transition objects
$\langle q_{i},\iota,q_{j} \rangle$, where a transition is made from state $q_{i}$ to state $q_{j}$
when encountering input $\iota$. In addition we specify the accepting states, which are all states
except the start state (white) and the bye state (pink).  This constraint also ensures that if a
skipper has finished his matches in flight $i$ (i.e.\ is in state $\END$) then he will also have
finished his matches in flight $i+1$ (and so on).

@@

        \begin{scope}
            \node [anchor=north west] at (7.5, 0.3) {
                \begin{minipage}{4.5cm}
                    \centering
                    \setlength{\tabcolsep}{3pt}
                    \begin{tabular}{cccc}
                        \toprule
                        Flight & \multicolumn{3}{c}{Matches} \\[2pt] \midrule
                        0 & (0,6) & (3,2) & (5,4) \\
                        1 & (0,2) & (4,1) & (6,5) \\
                        2 & (2,1) & (5,0) & (6,3) \\
                        3 & (1,5) & (3,0) & (4,6) \\
                        4 & (3,5) & (6,1) & (2,4) \\
                        5 & (1,3) & (0,4) & (5,2) \\
                        6 & (4,3) & (2,6) & (1,0) \\ \bottomrule
                    \end{tabular}
                \end{minipage}
            };
        \end{scope}

@@

We present three pairs of schedules. The first pair is for 8 skippers and 6 boats, one published by
ISAF, the other generated by us. Next, we present two new schedules, not appearing in the ISAF
manual, for 9 skippers and 6 boats and 9 skippers with 8 boats. The last pair is for 10 skippers and
6 boats, again ISAF published and generated by us. For each of our schedules, we used 1 day CPU time
for the first stage (minimising boat changes) then a further 2 days to minimise imbalance. Stages 3
and 4 (renaming and orientation) completed in less than a second.

@@

\paragraph{8 skippers and 6 boats:} We first analyse the ISAF schedule, Table~\ref{08-06} on the
left. Criterion~4 (skippers in last match in a flight cannot be first in next flight) is violated on
3 occasions (skipper 1 in flight 3, skipper 7 in flight 4, skipper 0 in flight 7). Criterion~5 (no
more than two consecutive port or starboard assignments) is violated for skipper 6 (flights 1 to 3)
and 7 (flights 7 to 9). Criterion~6 (imbalance) is violated for skippers 6 and 8. Criterion~12  (new
boats do not sail first) is violated in flight 9 for skipper 5. Finally, the schedule has 8 boat
changes. Our schedule, Table~\ref{08-06} on the right, respects all criteria and has 6 boat changes
and a maximum imbalance of 1.

\paragraph{9 skippers}: Two new schedules are presented in Table~\ref{09}, on the left for 6 boats
and on the right for 8 boats. Neither of these schedules appear in the ISAF Umpires' Manual. Both
schedules respect all criteria. For 6 boats there are 8 boat changes (no skipper making more than 1
boat change) and a maximum imbalance of 2. For 8 boats there are again 8 boat changes, no skipper
with more than 1 boat change, and perfect balance (each skipper 4 times on starboard side and 4 on
port side).

@@

\paragraph{10 skippers and 6 boats:} Table \ref{10-06} shows the ISAF schedule on the right for 10
skippers and 6 boats and on the right, our schedule.  The ISAF schedule violates criterion~5 (no more
than 2 consecutive port or starboard assignments) 10 times (twice for skipper 0, 1 and 6 and once
for skippers 3, 7, 8 and 9).  Criterion~12 (new boats do not sail first) is violated 7 times (in
flight 3 for skippers 6 and 2, flight 5 for skipper 8, flight 6 for skipper 7, flight 8 for skipper
1, flight 9 for skipper 4 and flight 12 for skipper 4 again). There are 22 boat changes, with the
minimum changes for a skipper being 1 and the maximum 3. Our schedule, right of Table \ref{10-06}
satisfies all criteria, has 12 boat changes with the minimum changes for a skipper being 0 and the
maximum 2, and a maximum imbalance of 1.

@@

@@

There are many redundant variables in our model. One of our first valid models was essentially that
shown in Figure~\ref{skipper5}, and did not use $\modMatch$ and $\match$ variables. Performance was
hopeless, struggling to produce a 7 skipper 6 boat schedule in days. The $\modMatch$ and $\match$
variables were added. This improved domain filtering, and a flattened $\match$ array was used as
decision variables. That is, the decisions where ``when do we schedule this match?''.  At this point
we had not yet produced a 9 skipper 6 boat schedule and we suspected that the combined criteria
might exclude such a schedule. A non-constraint model was developed by the fourth author, primarily
to attempt to prove that there was no schedule for 9 skippers with 6 boats. This program used a
backtracking search with decision variables being positions within flights, i.e.\ time slots being
assigned to matches. At the top of search, the matches in the first flight were set, and symmetry
breaking was used to reduce the number of legal second flights. A solution was found in seconds!
With this experience we added in the $\timeVar$ variables, channeled these into the existing model
and used these as decision variables, i.e.\ the question now was ``what match will we do now?''. We
also anchored the matches in the first flight. With these changes we began to produce schedules in
acceptable time. The model was then built upon, incrementally adding criteria. This took a
surprisingly short amount of time, sometimes minutes of coding to add in a new feature. The model
was then enhanced so that it would address optimisation rather than just satisfaction, again a
trivial programming task.

@@

So, why did we use constraint programming? The answer is obvious:  we are constraint programmers, or
to borrow Mark Twain's words ``To a man with a hammer, everything looks like a nail''. But actually,
constraint programming has been a good way to go. From an engineering perspective, it has allowed us
to prototype solutions quickly and to build solutions incrementally. There is also an advantage that
we might exploit in the future: we can now investigate the effect different criteria have on our
ability to produce schedules and how relaxation of those affect optimisation criteria. That is, the
constraint program might be used to design the next set of criteria for match race schedules, or
allow the event organiser to decide which criteria he may sacrifice to ensure a faster schedule with
less boat changes.

>>>>

This work describes a round robin competition format that arises from sailing, known as match
racing. Competitors, i.e.\ \emph{skippers}, compete against each other in a series of matches taking
place in rounds, known as \emph{flights}.  A match is composed of two skippers, with each skipper in
a boat on their own. Skippers in a match set off together, one on the port side, the other starboard,
and first home wins that match. Criteria published by the International Sailing Federation (ISAF)
dictate what makes a legal match race schedule and what should be optimised. This is a rich set of
criteria, 13 in all, and as far as we know, all published schedules have been produced by hand. This
is a daunting task. A close inspection of these schedules reveals that most are illegal, violating
many of the match race scheduling criteria, many schedules are missing, and some that are legal are
far from ISAF definition of optimality.

@@

Criteron~6 allows us to measure the \emph{balance} of a schedule. Perfect balance would be when a skipper has as many matches in first position in flights as in second position in flights, and so on. For example, if we had 9 skippers and 8 boats, each skipper would have 8 matches and could be in the first, second, third or fourth match of a flight. In perfect balance, each of the 8 skippers would appear twice in each position. Balance is one of our optimisation criteria.

Criterion~13 uses the term \emph{blanks} where conventionally we use the term \emph{bye}. That is, a blank is a bye and a bye is a flight where a skipper does not compete (and is ashore).

@@

We now describe the constraint models and processes used in each stage. In the text below we assume
there are $n$ skippers and $b$ boats, with $m = b/2$ matches in a flight. In total there are $t =
n(n-1)/2$ matches and $f = \lceil n(n-1)/m  \rceil$ flights. We use the schedule in
Table~\ref{tab1}, for 7 skippers and 6 boats, to help illustrate the constraint models.

@@

\paragraph{Modeling skippers:} The first thing we model, in the first half of
Figure~\ref{model:stage1}, is a skipper. Each skipper $\sigma$ has a temporal view of the schedule
in the array of constrained integer variables $\timeSlot[0]$ to $\timeSlot[{t-1}]$, and a state view
in the array of constraint integer variables $\state[0]$ to $\state[{f-1}]$. The state of this
skipper in flight $i$ is found in $\state[i]$, and this corresponds to time slots $\timeSlot[{i \cdot
m}]$ to $\timeSlot[{m(i+1)-1}]$.

@@

In \eqref{var:timeSlot} the array $\timeSlot$ is defined and in \eqref{var:state}, $\state$ is
defined. If variable $\timeSlot[{i}] = k$ and $k \geq 0$ then this skipper is in a match with
skipper~$k$ at time~$i$. Variable $\state[{i}]$ gives the state of this skipper in flight~$i$. The
cases in \eqref{C1} state that a skipper can be in a match in the first time slot in the flight, or in
the middle of the flight\footnote{i.e.\ not first and not last.}, or in the last time
slot of the flight. Alternatively, if all time slots in a flight are equal to $-1$ then the skipper
is in a bye (i.e. not in a match in this flight), and if all time slots in a flight are equal to $-2$ then the skipper has finished all matches.

We must then ensure that each skipper $\sigma$ is in a match with all other skippers $\{0, \nldots,
n-1\} \setminus \{\sigma\}$. This is \eqref{C2}, which is enforced by imposing a global
cardinality constraint \cite{globCard} on the array $\timeSlot$.

\paragraph{State transitions:} The $\state$ variables are then used to impose match race criterion~4
(if last in flight $i$ then not first in flight $i+1$), criterion~11 (if last in flight $i$ then not
in a bye in flight $i+1$) and criterion~12 (if flight $i$ is a bye then flight $i+1$ is not first).
These criteria are imposed by \eqref{C3}, which is a deterministic finite automaton (DFA) constraint
\cite{Pesant04}. The transitions for the DFA are shown pictorially in Figure~\ref{skipper5}: arcs
represent a possible transition from one state to another. The DFA is encoded as a set of transition objects
$\langle q_{i},\iota,q_{j} \rangle$, where a transition is made from state $q_{i}$ to state $q_{j}$
when encountering input $\iota$. In addition we specify the accepting states, which are all states
except the start state (white) and the bye state (pink).  This constraint also ensures that if a
skipper has finished all matches in flight $i$ (i.e.\ is in state $\END$) then that skipper will also have
finished all matches in flight $i+1$ (and so on).

@@

        \begin{scope}
            \node [anchor=north west] at (7.5, 0.3) {
                \begin{minipage}{4.5cm}
                    \centering
                    \setlength{\tabcolsep}{3pt}
                    \begin{tabular}{cccc}
                        \toprule
                        Flight & \multicolumn{3}{c}{Matches} \\ \midrule
                        0 & (0,6) & (3,2) & (5,4) \\
                        1 & (0,2) & (4,1) & (6,5) \\
                        2 & (2,1) & (5,0) & (6,3) \\
                        3 & (1,5) & (3,0) & (4,6) \\
                        4 & (3,5) & (6,1) & (2,4) \\
                        5 & (1,3) & (0,4) & (5,2) \\
                        6 & (4,3) & (2,6) & (1,0) \\ \bottomrule
                    \end{tabular}
                \end{minipage}
            };
        \end{scope}

@@

We present four pairs of schedules. The first pair is for 8 skippers and 6 boats, one published by
ISAF, the other generated by us. Next, we present two new schedules, not appearing in the ISAF
manual, for 9 skippers and 6 boats and 9 skippers with 8 boats, then 10 skippers and
6 boats, again ISAF published and generated by us. Finalyy, we show what is a relatively easy problem, 8 skippers and 8 boats. 
For each of our schedules (with the exception of the "easy-8" problem), we used 1 day CPU time for the first stage (minimising boat changes) then a further 2 days to minimise imbalance. Stages 3 and 4 (renaming and orientation) completed in less than a second.

@@

\paragraph{8 skippers and 6 boats:} We first analyse the ISAF schedule, Table~\ref{08-06} on the
left. Criterion~4 (skippers in last match in a flight cannot be first in next flight) is violated on
3 occasions (skipper 1 in flight 3, skipper 7 in flight 4, skipper 0 in flight 7). Criterion~5 (no
more than two consecutive port or starboard assignments) is violated for skipper 6 (flights 1 to 3)
and skipper 7 (flights 7 to 9). Criterion~6 (imbalance) is violated for skippers 6 and 8. Criterion~12  (new
boats do not sail first) is violated in flight 9 for skipper 5. Finally, the schedule has 8 boat
changes. Our schedule, Table~\ref{08-06} on the right, respects all criteria and has 6 boat changes
and a maximum imbalance of 1.

\paragraph{9 skippers}: Two new schedules are presented in Table~\ref{09}, on the left for 6 boats
and on the right for 8 boats. Neither of these schedules appear in the ISAF Umpires' Manual. Both
schedules respect all criteria. For 6 boats there are 8 boat changes (no skipper making more than 1
boat change) and a maximum imbalance of 2. For 8 boats there are again 8 boat changes, no skipper
with more than 1 boat change, and each skipper 4 times on starboard side and 4 on port side.

@@

\paragraph{10 skippers and 6 boats:} Table \ref{10-06} shows the ISAF schedule on the right for 10
skippers and 6 boats and on the right, our schedule.  The ISAF schedule violates criterion~5 (no more
than 2 consecutive port or starboard assignments) ten times (twice for skippers 0, 1 and 6 and once
for skippers 3, 7, 8 and 9).  Criterion~12 (new boats do not sail first) is violated 7 times (in
flight 3 for skippers 6 and 2, flight 5 for skipper 8, flight 6 for skipper 7, flight 8 for skipper
1, flight 9 for skipper 4 and flight 12 for skipper 4 again). There are 22 boat changes, with the
minimum changes for a skipper being 1 and the maximum 3. Our schedule, right of Table \ref{10-06},
satisfies all criteria, has 12 boat changes with the minimum changes for a skipper being 0 and the
maximum 2, and a maximum imbalance of 1.

@@

\begin{center}
\begin{figure}[h]
\begin{minipage}[t]{0.48\textwidth}
    \centering
\begin{tabular}{ccccc}
        \toprule
        Flight & \multicolumn{4}{c}{Matches} \\ \midrule
0 & (5,2) & (4,3) & (1,6) & (0,7) \\
1 & (2,4) & (0,6) & (1,7) & (5,3) \\
2 & (0,5) & (7,2) & (6,3) & (4,1) \\
3 & (7,3) & (5,1) & (4,0) & (6,2) \\
4 & (3,0) & (4,7) & (6,5) & (2,1) \\
5 & (6,4) & (7,5) & (2,0) & (3,1) \\
6 & (7,6) & (5,4) & (3,2) & (1,0) \\ \bottomrule
    \end{tabular}
\label{08-08a}
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
    \centering
\begin{tabular}{ccccc}
        \toprule
        Flight & \multicolumn{4}{c}{Matches} \\ \midrule
0 & (4,3) & (2,0) & (1,5) & (7,6) \\
1 & (3,2) & (0,4) & (6,1) & (5,7) \\ 
2 & (4,1) & (7,3) & (5,2) & (0,6) \\
3 & (1,7) & (6,5) & (3,0) & (2,4) \\ 
4 & (0,5) & (4,6) & (7,2) & (3,1) \\
5 & (7,0) & (2,1) & (6,3) & (5,4) \\ 
6 & (6,2) & (5,3) & (4,7) & (1,0) \\ \bottomrule
    \end{tabular}
\label{08-08b}
\end{minipage}
\caption{Schedules for 8 skippers and 8 boats. On the left, the published ISAF schedule (illegal),
and on the right, our schedule.}
\label{08-08}
\end{figure}
\end{center}

\paragraph{8 skippers and 8 boats (easy-8):} Table \ref{08-08} shows the ISAF schedule on the right for 8
skippers and 8 boats and on the right, our schedule.  The ISAF schedule violates criterion~4 (last match in a flight cannot be first in next flight) for skipper 5 in flights 1 and 2. Also, criterion 5 (no skipper should have more than 2 consecutive port or starboard assignments) is violated for skippers 0, 1, 3, 4 and 6. Furthermore, skipper 1 appears in the last match of four flights resulting in significant imbalance. Our schedule, on the right, satisfies all criteria, is optimal (maximum imbalance of 1) and took less than 10 seconds in total (all stages) to produce.

@@

There are many redundant variables in our model. One of our first valid models was essentially that
shown in Figure~\ref{skipper5}, and did not use $\modMatch$ and $\match$ variables. Performance was
hopeless, struggling to produce a 7 skipper 6 boat schedule in days. The $\modMatch$ and $\match$
variables were added. This improved domain filtering, and a flattened $\match$ array was used as
decision variables. That is, the decisions where ``when do we schedule this match?''.  At this point
we had not yet produced a 9 skipper 6 boat schedule and we suspected that the combined criteria
might exclude such a schedule. A non-constraint model was developed, primarily
to attempt to prove that there was no schedule for 9 skippers with 6 boats. This program used a
backtracking search with decision variables being positions within flights, i.e.\ time slots being
assigned to matches. At the top of search, the matches in the first flight were set, and symmetry
breaking was used to reduce the number of legal second flights. A solution was found in seconds!
With this experience we added in the $\timeVar$ variables, channeled these into the existing model
and used these as decision variables, i.e.\ the question now was ``what match will we do now?''. We
also anchored the matches in the first flight. With these changes we began to produce schedules in
acceptable time. The model was then built upon, incrementally adding criteria. This took a
surprisingly short amount of time, sometimes minutes of coding to add in a new feature. The model
was then enhanced so that it would address optimisation rather than just satisfaction, again a
trivial programming task.

@@

So, why did we use constraint programming? The answer is obvious:  we are constraint programmers, or
to borrow Mark Twain's words ``To a man with a hammer, everything looks like a nail''. But actually,
constraint programming has been a good way to go. From an engineering perspective, it has allowed us
to prototype solutions quickly and to build solutions incrementally. There is also an advantage that
we might exploit in the future: we can now investigate the effect different criteria have on our
ability to produce schedules and how relaxation of those affect optimisation criteria. That is, the
constraint program might be used to design the next set of criteria for match race schedules, or
allow the event organiser to decide which criteria to sacrifice to ensure a faster schedule with
less boat changes.
